Weighted Model Counting as a Special Case of Polyadic Measure Algebras

### 2020-03-09 paper.tex
Proof of the obvious result: part 1

\Crefname{property}{Property}{Properties}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{nWMC}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\End}{End}

@@

@@

\begin{theorem} \label{thm:representation}
  Let $\mathbf{B}$ be a finite Boolean algebra. Then every $a \in \mathbf{B}
  \setminus \{ 0 \}$ can be uniquely expressed as $a = \bigvee_{i \in I} a_i$,
  where $\{ a_i \}_{i \in I}$ is the set of atoms in $\mathbf{B}$ such that $a_i
  \le a$ for all $i \in I$.
\end{theorem}
\begin{proof}
  A simple consequence of the theorem that every finite Boolean algebra is
  isomorphic to a field of subsets of a set, where the cardinality of the set is
  equal to the number of atoms in the Boolean algebra.
\end{proof}

@@

\todo[inline]{Allow weight to be zero}
\begin{definition}
  Let $\mathbf{B}$ be a finite Boolean algebra, and let $M \subseteq \mathbf{B}$
  be its set of atoms. Let $L \subseteq \mathbf{B}$ be such that every atom $m
  \in M$ can be uniquely expressed as $m = \bigwedge_{i \in I} l_i$ for some $\{
  l_i \}_{i \in I} \subseteq L$, and let $w : L \to \mathbb{R}_{>0}$ be
  arbitrary. The \emph{weighted model count} $\WMC : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

\todo[inline,caption={}]{The big TODO list
  \begin{itemize}
  \item Condition for a measurable Boolean algebra to be representable as WMC.
    Note: if all atoms are in $L$, then WMC can define any distribution.
  \item Extension to infinite (atomic?) Boolean algebras.
  \item How many extra variable do you need to add to make any probability
    distribution representable using WMC?
  \item Abstraction refinements as homomorphisms.
  \item Definition of a measure-preserving homomorphism from Jech's set theory
    book.
  \item A Boolean algebra is approximable if its Stone space is approximable.
  \end{itemize}
}

>>>>

\Crefname{property}{Property}{Properties}
\Crefname{condition}{Condition}{Conditions}
\creflabelformat{condition}{#2(#1)#3}

\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{nWMC}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\End}{End}

@@

% Big advantage: using algebras we can represent an incomplete state of
% knowledge, e.g., we know m(a \/ b), but not m(a) or m(b). In infinite BAs, one
% can have measure undefined for some elements.

@@

% TODO: we don't have to treat 0 as an exception we state that \/ of an empty set is 0 (and /\ of an empty set is 1).
% TODO: use bigvee and bigwedge directly with sets.
\begin{theorem} \label{thm:representation}
  Let $\mathbf{B}$ be a finite Boolean algebra. Then every $x \in \mathbf{B}
  \setminus \{ 0 \}$ can be uniquely expressed as
  \[
  x = \bigvee_{\text{atoms } a \le x} a.
  \]
\end{theorem}
\begin{proof}
  A simple consequence of the theorem that every finite Boolean algebra is
  isomorphic to a field of subsets of a set, where the cardinality of the set is
  equal to the number of atoms in the Boolean algebra.
\end{proof}

@@

\todo[inline]{Allow weight to be zero}
\begin{definition} \label{def:wmc}
  Let $\mathbf{B}$ be a finite Boolean algebra, and let $M \subseteq \mathbf{B}$
  be its set of atoms. Let $L \subseteq \mathbf{B}$ be such that every atom $m
  \in M$ can be uniquely expressed as $m = \bigwedge_{i \in I} l_i$ for some $\{
  l_i \}_{i \in I} \subseteq L$, and let $w : L \to \mathbb{R}_{>0}$ be
  arbitrary. The \emph{weighted model count} $\WMC : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

% TODO: maybe I should gives this kind of a BA a name? A synonym of 'complete',
% perhaps.
% TODO: a special case for weight=0.
\begin{proposition}
  Let $\mathbf{B}$ be a finite measure algebra with measure $m : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$. Let $L \subset \mathbf{B}$ be defined as
  \[
  L = \{ l_i \mid i \in [n] \} \cup \{ \neg l_i \mid i \in [n] \}
  \]
  for some $n \in \mathbb{N}$. Finally, assume that $\mathbf{B}$ has $2^n$
  atoms, where each atom $a \in \mathbf{B}$ is an infimum
  \[
    a = \bigwedge_{i=1}^n a_i
  \]
  such that $a_i \in \{ l_i, \neg l_i \}$ for $i \in [n]$. Then there exists a
  weight function $w : L \to \mathbb{R}_{>0}$ that makes $m$ a WMC measure if
  and only if
  \begin{equation} \label{eq:wmccondition}
  m(l \land l') = m(l)m(l')
  \end{equation}
  for all distinct $l, l' \in L$ such that $l \ne \neg l'$.
\end{proposition}

\begin{remark}
  Note that if $n = 1$, then \cref{eq:wmccondition} is vacuously satisfied and
  so any valid measure can be expressed as WMC.
\end{remark}

\begin{proof}
  Let us begin with the `if' part of the statement. Let $w : L \to
  \mathbb{R}_{>0}$ be defined by
  \begin{equation} \label{eq:assumption}
    w(l) = m(l)
  \end{equation}
  for all $l \in L$. We are going
  to show that $\nWMC = m$. First, note that $\nWMC(0) = 0 = m(0)$ by the
  definitions of both $\nWMC$ and $m$. Second, let
  \begin{equation} \label{eq:def_of_a}
    a = \bigwedge_{i=1}^n a_i
  \end{equation}
  be an atom in $\mathbf{B}$ such that $a_i \in \{ l_i, \neg l_i \}$ for all $i
  \in [n]$. Then
  \[
    \nWMC(a) = \frac{\WMC(a)}{\WMC(1)} = \frac{1}{\WMC(1)} \prod_{i=1}^n w(a_i)
    = \frac{1}{\WMC(1)} \prod_{i=1}^n m(a_i) = \frac{1}{\WMC(1)} m \left(
      \bigwedge_{i=1}^n a_i \right) = \frac{m(a)}{\WMC(1)}
  \]
  by \cref{def:wmc,eq:assumption,eq:wmccondition,eq:def_of_a}. Now we just need
  to show that $\WMC(1) = 1$. Indeed,
  \begin{align*}
    \WMC(1) &= \sum_{\text{atoms } a \in \mathbf{B}} \WMC(a) = \sum_{\text{atoms
      } a \in \mathbf{B}} \prod_{i=1}^n w(a_i) = \sum_{\text{atoms } a \in
      \mathbf{B}} \prod_{i=1}^n m(a_i) \\
    &= \sum_{\text{atoms } a \in
      \mathbf{B}} m \left( \bigwedge_{i=1}^n a_i \right) = \sum_{\text{atoms } a
      \in \mathbf{B}} m(a) = m \left( \bigvee_{\text{atoms } a \in \mathbf{B}}
    \right) = m(1) = 1.
  \end{align*}
  Finally, note that if $\nWMC$ and $m$ agree on all atoms, then they must also
  agree on all other non-zero elements of the Boolean algebra.

  % TODO: the other direction
\end{proof}

\todo[inline,caption={}]{The big TODO list
  \begin{itemize}
  \item Extension to infinite (atomic?) Boolean algebras.
  \item How many extra variable do you need to add to make any probability
    distribution representable using WMC?
  \item Abstraction refinements as homomorphisms.
  \item Definition of a measure-preserving homomorphism from Jech's set theory
    book.
  \item A Boolean algebra is approximable if its Stone space is approximable.
  \end{itemize}
}
